### 2014-05-14 writeups/context-dependence.tex
writeup edits

Equipped with marginal transition probabilities from length-$m+2\ell$ patterns to length-$m$ patterns
that correctly account for local dependencies,
we can now consider inference procedures.

Now we know how to compute the approximate (marginal) probability that a particular subsequence changes into another subsequence.
Let $N(x)$ be the number of occurrences of a pattern $x$ in $X(0)$.
For patterns $x$ and $y$, with both fringe widths equal to $\ell$ (so that $|x|=|y|+2\ell$),
let $N_\ell(x,y)$ count these events, respectively be the number of times $x$ appears in $X(0)$,
and how many times $x$ appears in $X(0)$ while $y$ appears at the same place in $X(t)$, offset by $\ell$.
For instance, $N_3(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  C  & A & C & T & C & A & C \\
 $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
\end{tabular}
\end{center}
where $\cdot$ represents any base.

\newcommand{\lyl}{\ell,|y|,\ell}

Now we will now calculate the expectation and covariances of $N_\ell(x,y)$.
From the definitions above,
\[
\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y).
\]
To compute covariances, define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$.
\todo[inline]{
In the below, are you are using an equivalence
\[
\cov[M_1(x,y),M_{k+1}(x,y)] = \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2
\]
How do you justify this? I don't think that $M_1(x,y)$ is equal to $p_{\lyl}(x,y)$: the former is defined in terms of observed events, whereas the latter is defined starting with "if we have a sequence $x$" without any reference to the frequency of $x$.
For example, what if $x$ doesn't appear in $X(0)$?
I was with you on $\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y)$.
}

\begin{align}
  \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^{n-1} \cov[M_1(x,y),M_{k+1}(x,y)] \\
  &\approx n p_{\lyl}(x,y)(1-p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2 \right) \\
  &= n p_{\lyl}(x,y)(1-(1+2m+4\ell)p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
\end{align}
The term $\E[M_1(x,y) M_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
\begin{align}
  \begin{split}
    \cov[N(x_1,y_1),N(x_2,y_2)]
    & \approx - n (2m+4\ell) p_{\lyl}(x_1,y_1)p_{\lyl}(x_2,y_2) \\
    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
  \end{split}
\end{align}

@@

The covariance calculations show that
if $p_{\lyl}(x,y)$ is small enough (i.e.\ the pattern is not common),
then $N(x,y)$ is close to Poisson.
Using this observation, we can estimate parameters by composite likelihood,
using all $N(x,y)$ as data.
Maximum likelihood then gets an asymptotically unbiased estimate of the parameters.
However, since they are not independent, to get confidence intervals, we need to either simulate, or do something different.

@@

In fact, the approximation improves exponentially as $\ell$ and $m$ increase --
here is a simple argument.
Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
The transition rate at site $i$ at time $s$ is determined if we know $X_{i-R}^{(2R+1)}(s)$;
this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.
These will depend in turn on the outcome of previous possible changes.
For each site $i$ and times $s_1 < s_2$, define the set of states $H(i,s_1,s_2)$
so that if one knows the possible change points $s_{ij}$,
then changes outside of $H(i,s_1,s_2)$ do not affect the distribution of $X_i(s_2)$.
Concretely, define $H(i,s_1,s_2)$ by the property that $i \in H(i,s_1,s_2)$, and
\begin{gather}
  \text{if } s_1 < s_{jk} \le s_2 \text{ for some } i-R-1 \le j \le i+R, \text{ then } H(j,s_1,s_{jk}) \subseteq H(i,s_1,s_2) .
\end{gather}
In other words, $H(i,s_1,s_2)$ is the union of the $R$-neighborhoods of $i$ and all possible changes occurring in $H(i,s_1,s_2)$ between $s_1$ and $s_2$.
For a set of states $I$, let $H(I,s_1,s_2) = \bigcup_{i\in I} H(i,s_1,s_2)$.
Formally, the state $X_i(s_2)$ is conditionally independent of $\{X_j(s_1) \st j \notin H(i,s_1,s_2)\}$.
Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does note expand beyond the flanking context.
Now, MAKE THIS EASIER TO READ
\begin{multline}
  \P\{ X_i^{(m)}(t) = y \mid X(0) \}
  = \P\left\{ H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
   \qquad \qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
     \qquad {} + \P\left\{ H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\}  \\
   \qquad\qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\} ,
\end{multline}
and so
\begin{align} \label{eqn:prob_approx}
  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\lyl}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
\end{align}

>>>>

Since we cannot compute reasonably the full likelihood for the data,
we will need to choose appropriate summary statistics
that contain as much information about the process as possible,
and have a tractable expression for their marginal likelihood.
Assume here that $\ell$ has been chosen to be sufficiently large that approximation \eqref{eqn:window_approx} holds exactly.
If $i_1 < \ldots < i_{n(x)}$ are a set of nonoverlapping sites where the initial sequence matches $x$
-- i.e.\ $X_{i_k}^{(2\ell+m)}(0)=x$ and $i_{k+1} > i_k + 2\ell+m$ for all $k$ --
then the counts $N(x,y)=\#\{ i_k \st X_{i_k+\ell}^{(m)}(t)=y\}$ (the number of these sites that match the smaller pattern $y$ at time $t$)
are multinomial, with probabilities $\{p_{\ell,m}(x,y)\st y \in \calS^m\}$.
For instance, if $\ell=3$ then $N(\text{ACTCAC}, \text{CGC})$ counts the number of occurrences of
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  C  & A & C & T & C & A & C \\
 $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
\end{tabular}
\end{center}
where $\cdot$ represents any base.
More generally, if we partition the sequence into windows of length $2\ell+m$,
and let $N(x,y)$ be the number of times we see a window in which $x$ matches $X(0)$ and $y$ matches $X(t)$,
then these counts are independently multinomial across each $x$.
% so the marginal likelihood of the statistics $N(x,y)$ is
% \begin{align}
%     \prod_{x \in \calS^{2\ell+m}} N(x)! \prod_{y \in \calS^m} \frac{p_{\ell,m}(x,y)^{N(x,y)}}{N(x,y)!} ,
% \end{align}
% where $N(x) = \sum_y N(x,y)$.

Of course, a drawback to this approach is that we throw away a good fraction of the data.
In fact, we do not even look at a fraction $2\ell/(2\ell+m)$ of the second sequence, $X(t)$,
that lies in the fringes of each window.
As before, let $N(x)$ be the number of occurrences of a pattern $x$ in $X(0)$,
and $N(x,y)$ the number of times $x$ is seen juxtaposed with $y$ in $X(t)$,
but this time, count over all positions,
so that e.g.\ $N(x) = \#\{ 1 \le i \le n - 2\ell - m + 1 \st X_i^{(m+2\ell)}(0)=x\}$.

\peter{Ooops, yes.  Previously, I was semi-inconsistently thinking of the initial sequence $X(0)$ as random as well. ,
    which is why I was talking about Poisson instead of Multinomial counts.}

Now we will now calculate the expectation and covariances of $N(x,y)$.
\peter{The following is Not Right, but also we don't really use it at the moment.}
If we assume that \eqref{eqn:window_approx} is exact, then
\[
    \E [N(x,y)] = N(x) \, p_{\ell,m}(x,y).
\]
To compute covariances, define $\chi_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
so that $N(x,y) = \sum_{i=1}^n \chi_i(x,y)$.
Overloading notation, also define $\chi_i(x)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$,
\begin{align}
  \var[N(x,y)] &= \sum_{i=1}^n \var[\chi_i(x,y)] + \sum_{i\neq j} \cov[\chi_i(x,y), \chi_j(x,y)] \\
               &\approx \sum_{i=1}^n \chi_i(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) 
            + 2 \sum_{i=1}^{n-2\ell-m} \sum_{j=i+1}^{i+2\ell+m-1} \cov[ \chi_i(x,y) \chi_j(x,y) ] \\
            &= N(x) p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) + 
\end{align}
The term $\E[\chi_1(x,y) \chi_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
\begin{align}
  \begin{split}
    \cov[N(x_1,y_1),N(x_2,y_2)]
    & \approx - n (2m+4\ell) p_{\ell,m}(x_1,y_1)p_{\ell,m}(x_2,y_2) \\
    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[\chi_1(x_1,y_1) \chi_{k+1}(x_2,y_2) + \chi_1(x_2,y_2) \chi_{k+1}(x_1,y_1)] .
  \end{split}
\end{align}

@@

The covariance calculations show that
if $p_{\ell,m}(x,y)$ is small enough (i.e.\ the pattern is not common),
then $N(x,y)$ is close to Poisson.
Using this observation, we can estimate parameters by composite likelihood,
using all $N(x,y)$ as data.
Maximum likelihood then gets an asymptotically unbiased estimate of the parameters.
However, since they are not independent, to get confidence intervals, we need to either simulate, or do something different.

@@

In fact, the approximation improves exponentially as $\ell$ and $m$ increase --
here is a simple argument.
Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
The transition rate at site $i$ at time $s$ is determined if we know $X_{i-R}^{(2R+1)}(s)$;
this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.
These will depend in turn on the outcome of previous possible changes.
For each site $i$ and times $s_1 < s_2$, define the set of states $H(i,s_1,s_2)$
so that if one knows the possible change points $s_{ij}$,
then changes outside of $H(i,s_1,s_2)$ do not affect the distribution of $X_i(s_2)$.
Concretely, define $H(i,s_1,s_2)$ by the property that $i \in H(i,s_1,s_2)$, and
\begin{gather}
  \text{if } s_1 < s_{jk} \le s_2 \text{ for some } i-R-1 \le j \le i+R, \text{ then } H(j,s_1,s_{jk}) \subseteq H(i,s_1,s_2) .
\end{gather}
In other words, $H(i,s_1,s_2)$ is the union of the $R$-neighborhoods of $i$ and all possible changes occurring in $H(i,s_1,s_2)$ between $s_1$ and $s_2$.
For a set of states $I$, let $H(I,s_1,s_2) = \bigcup_{i\in I} H(i,s_1,s_2)$.
Formally, the state $X_i(s_2)$ is conditionally independent of $\{X_j(s_1) \st j \notin H(i,s_1,s_2)\}$.
Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does note expand beyond the flanking context.
Now, MAKE THIS EASIER TO READ
\begin{multline}
  \P\{ X_i^{(m)}(t) = y \mid X(0) \}
  = \P\left\{ H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
   \qquad \qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
     \qquad {} + \P\left\{ H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\}  \\
   \qquad\qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\} ,
\end{multline}
and so
\begin{align} \label{eqn:prob_approx}
  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
\end{align}
